---
title: "Assignment 5: Data Visualization"
author: "Student Name"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 Project setup
##Load packages
library(tidyverse);library(lubridate);library(here);library(cowplot)
## Check working directory
here()
##Load datasets
NTL_tidy <- read.csv(
  here('Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv'),
  stringsAsFactors = T)
Trap_data <- read.csv(
  here('Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv'),
  stringsAsFactors = T)

#2 Check/convert date columns
NTL_tidy$sampledate <- ymd(NTL_tidy$sampledate)
Trap_data$collectDate <- ymd(Trap_data$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3 Create my theme
my_theme = theme_bw() + 
  theme(
    plot.background = element_rect(
      fill = 'Light gray',
      color = 'Black'
    ),
    axis.line = element_line(
      linewidth = 1,
      color='Blue'
    ),
    axis.text = element_text(
      family='serif'
    ),
    plot.title = element_text(
      face = 'bold',
      color='Blue',
      family='serif'
    )
  )

#Set my theme to the default
theme_set(my_theme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 Plot total P vs PO4
ggplot(NTL_tidy,aes(x=po4,y=tp_ug,color=lakename)) + 
  geom_point() + 
  geom_smooth(method='lm')+
  xlim(0,50)+
  labs(
    x = expression(paste('Phosphate (',mu,'/g')),
    y = expression(paste('Total Phosphorus (',mu,'/g)')),
    title = 'Q4: Plot of Total Phosphorus vs Phosphate',
    color='Lake'
    ) 

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 Boxlots
base_plt <- NTL_tidy %>% 
  mutate(the_month = factor(month,levels = 1:12, labels = month.abb)) %>% 
  ggplot(aes(x=the_month,color=lakename)) #+
  scale_x_discrete(labels=month.abb) 

plt1 <- base_plt+
  geom_boxplot(aes(y=temperature_C),show.legend = T) +
  labs(
    x='',
    y='Temperature (C)',
    title='Temperature'
  ) + 
  theme(
    legend.position = "top"
  )

plt2 <- base_plt+
  geom_boxplot(aes(y=tp_ug),show.legend = F) +
  labs(
    x='',
    y='Total P',
    title='Total P'
  )

plt3 <- base_plt+
  geom_boxplot(aes(y=tn_ug),show.legend = F) +
  labs(
    x='Month',
    y='Total N',
    title='Total N',
    color='Lake'
  ) 

plot_grid(plt1,plt2,plt3,
          nrow = 3, 
          align = 'h', 
          rel_heights = c(1.3,1,1))

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
Trap_data %>% 
  filter(functionalGroup == 'Needles') %>% 
  ggplot(aes(x=collectDate,y=dryMass,color=nlcdClass)) + 
  geom_point() 

#7
Trap_data %>% 
  filter(functionalGroup == 'Needles') %>% 
  ggplot(aes(x=collectDate,y=dryMass)) + 
  geom_point() +
  facet_wrap(vars(nlcdClass))

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: