---
title: "Assignment 10: Data Scraping"
author: "Student Name"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 Import packages & check path
library(tidyverse);library(here);library(rvest)
here()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 create the web page object
the_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022'
webpage <- read_html(the_url)

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 Scrape items from the web page object

#Water system name
water.system.name <- webpage %>% 
  html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% 
  html_text()
  
#PWSID
PWSID <- webpage %>% 
  html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% 
  html_text()
  
#Ownership
ownership <-  webpage %>% 
  html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% 
  html_text()

#Max withdrawal data
max.withdrawals.mgd <-  webpage %>% 
  html_nodes('th~ td+ td') %>% 
  html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 Combine items into a dataframe
the_df <- data.frame(
  "water.system.name" = rep(water.system.name,12),
  "PWSID" = rep(PWSID,12),
  "ownership" = rep(ownership,12),
  "max.withdrawals.MGD" = as.numeric(max.withdrawals.mgd),
  "month" = c(1,5,9,2,6,10,3,7,11,4,8,12), #! Months need to be in order scraped
  "year" = 2022
) %>% 
  mutate(the_date = my(paste(month,year))) 

#5 Plot
ggplot(the_df,aes(x=the_date,y=max.withdrawals.MGD)) + 
  geom_line() + 
  geom_smooth(method = 'loess')

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. Create a scraping function for a given PWSID and Year
scrape.it <- function(the_pwsid, the_year){
  #Construct the url, inserting the PWSID and year
  the_url <- paste0(
    'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',
    the_pwsid,
    '&year=',
    the_year)
  #Create the web page object
  webpage <- read_html(the_url)
  
  #Scrape the data (as above); no need to scrape the PWSID since it's an input
  water.system.name <- webpage %>% 
    html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% 
    html_text()

  ownership <-  webpage %>% 
    html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% 
    html_text()
  
  max.withdrawals.mgd <-  webpage %>% 
    html_nodes('th~ td+ td') %>% 
    html_text()
  
  avg.withdrawals.mgd <-  webpage %>% 
    html_nodes('.fancy-table:nth-child(31) th+ td') %>% 
    html_text()
  
  #Construct a dataframe from scraped data
  the_df <- data.frame(
    "water.system.name" = rep(water.system.name,12),
    "PWSID" = rep(the_pwsid,12),
    "ownership" = rep(ownership,12),
    "max.withdrawals.MGD" = as.numeric(max.withdrawals.mgd),
    "month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
    "year" = the_year
  ) %>% 
    mutate(the_date = my(paste(month,the_year)))
  
  #Return the constructed dataframe
  return(the_df)
    
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 Apply the function to PWSID 03-32-010 in year 2015
durham_df <- scrape.it('03-32-010',2015)

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
#Apply the function to PWSID 01-11-010 in year 2015
asheville_df <- scrape.it('01-11-010',2015)

#Combine with previously scraped dataframe
combined_df <- rbind(durham_df, asheville_df)

#Plot, coloring line by PWSID
ggplot(combined_df,aes(x=the_date,y=max.withdrawals.MGD,color=PWSID)) + 
  geom_line() + 
  labs(
    x = "",
    y = "Max Withdrawals (MGD)",
    color = 'PWSID',
    title = "Maximum monthly water withrawals, 2015"
  ) + 
  scale_x_date(
    date_breaks = "1 months",
    date_labels = "%b"
  )


```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
the_years = seq(2010,2022)
the_site = rep('01-11-010',length(the_years))
asheville_multiyear <- map2(the_site,the_years,scrape.it) %>% bind_rows()

ggplot(asheville_multiyear,aes(x=the_date,y=max.withdrawals.MGD)) +
  geom_line(color='blue') +
  scale_x_date(
    date_breaks = "1 year",
    minor_breaks = NULL,
    date_labels = '%Y'
  ) + 
  geom_smooth(method = 'loess')+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) +
  labs(
    title = 'Max daily withdrawal',
    x = NULL,
    y = "Max daily withdrawal (MGD)")
```

> Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
